Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-17 forecastflows.tex
Small changes in last section

\subsection{Adopting the workflow}
Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training. Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models. In ecological training today researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test). Such training leaves a large number of ecologists adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is. % jdavies15april: this may be true, but I am not sure we want to broadly disparage ecologists

Instead, ecologists trained in a limited set of tests are expected to collaborate with others when they need more complex models, who were trained more in model development (though often for highly specific applications, such as wildlife population estimates, where they may rarely develop entirely new generative models). These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models. % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental training in ecology has overly-siloed these groups and prevented more rapid progress.

Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science. % jdavies15april: this feels a little OTT
Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in our workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them. Through this and retrodictive checks most ecologists would more easily think through what parameters are most critical to their question and or aim (e.g., management) and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting. 
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches to better forecasts or something. % jdavies15april: and parameter estimation to improve forecasts.

>>>>

\subsection{Adopting the workflow}

Advancing ecology to where most researchers use models built more flexibly from ecological theory and insights applied to their ecological systems will not happen rapidly without a major shift in training. Much of ecology still divides the world into training for those focused on several groups: those who gather data and learn a limited set of pre-built models versus those who develop more complex models. In ecological training today researchers who conduct field and lab studies often learn a highly limited set of particular tests matched to particular experiment designs and simple information on their variable types (e.g., categorical $x$ and $y$ leads to using a chi square test). 
% Such training leaves a large number of ecologists adrift later when asked to simulate any sort of simple model because they were never taught what a model fundamentally is. % jdavies15april: this may be true, but I am not sure we want to broadly disparage ecologists

Instead, ecologists trained in a limited set of tests are expected to collaborate with others when they need more complex models, who were trained more in model development (though often for highly specific applications, such as wildlife population estimates, where they may rarely develop entirely new generative models). These two groups further differ from process-based modelers, who often train in physical and ecophysiological processes and how to abstract them into mainly deterministic models. % process-based modelers---often with a physical or ecophysiological background---mostly focus on simulating some mechanisms 
Few of these groups have integrated data simulation into their statistical or scientific workflows, which is generally reserved as a form of training needed mainly by those specializing in theoretical ecology, who often solve analytical equations but rarely link to empirical data. While specialization is valuable, we argue the fundamental % training 
separation in ecology has overly-siloed these groups and prevented more rapid progress.

% Training all ecologists in our proposed workflow would break down barriers between these different groups with likely major benefits for science. % jdavies15april: this feels a little OTT
Instead of training some ecologists extensively in experimental design and tests that may match certain designs (though rarely do for ecological data, CITES), training in the workflow would focus on learning to generate questions and then models, and then to simulate data from them. This would mean training all ecologists to link the ecological processes they study with the mathematical models that may describe them. Through this and retrodictive checks most ecologists would more easily think through what parameters are most critical to their question and or aim (e.g., management) and also gain a much stronger connection to the level of uncertainty in many of ecological estimates. Empirical ecologists would be more likely to recognize critical gaps in current models fit by those specializing in ecological modeling and help advance those models. Process-based modelers may start a new generation of simpler models that are more tractable to theoretical ecologists, who may suddenly see bridges from their work to empirical data and forecasting. 
% Perhaps the only drawback would be to those focused on statistical models, whose gate-keeping would suddenly fail them and they would have to do more that just say how amazing they are... 
Those focused on learning complex statistical models may find many of the field now share their excitement and interest in more generative models, and could focus on adapting approaches for better parameter estimation and improved forecasts.
